Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-07-18 WholePlotTrajectories2.tex
Final_V1

% User-adder preamble
\usepackage{textcomp} \DeclareUnicodeCharacter{B0}{\textdegree}
\usepackage{tabu} \usepackage{caption}
\captionsetup{justification = justified}
\renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
\hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}

@@

\Abstract{
Anticipate the impact of the global change context on tropical forests
is now urgent. This calls to highlight the taxonomic and functional
facets of tree community post-disturbance trajectories and the
underlying processes. It also calls to clarify the link between
biodiversity and disturbance, specifically the scope of the Intermediate
Disturbance Hypothesis (IDH) that is debated in tropical forests. In
this study, we analyzed the tree community trajectories following a
disturbance gradient in a Neotropical forest over 30 years. We
considered community taxonomic and functional diversity, composition and
redundancy trajectories. We based on the annual botanical inventories of
75 ha of a Neotropical forest and on large trait datasets comprising 7
leaf, stem and life-history traits. We highlighted a decoupling between
taxonomic trajectories, that differed among communities, and functional
trajectories, remained similar and convergent. We explained this
decoupling by the variations in community functional redundancy
mitigating the functional impact of disturbance. The humped-shaped
taxonomic trajectories which amplitude varied with disturbance intensity
validated the IDH (Intermediate Disturbance Hypothesis). The IDH was
however dismissed for functional diversity. Although consistent, the
recovery of community composition, diversity and redundancy remained
unachieved after 30 years. This acknowledged the need of decades-long
cycles without disturbance to ensure a complete recovery, and questioned
tropical forest community resilience after repeated disturbances.
}

@@

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events, stating that community diversity peaks
at intermediate level of disturbance \citep{Connell1978}. The theory
assumes that the environmental fluctuations following disturbance foster
both competitively superior species and fast colonizers and thus prevent
competitive exclusion \citep{Shea2004, Pulsford2016}. In tropical
forests, however, observations of the IDH often diverge from theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. In tropical forests the
processes underlying the IDH might indeed be complicated by the huge
diversity found in tropical tree communities and the variety of
ecological processes that may underlie the IDH (colonization,
facilitation, coexistence, etc)
\citep{Lindenmayer2012, Garcia_florez2017}. It is then crucial to test
the IDH in this context and determine its scope in time, using ecosystem
time-explicit trajectories \citep{Sheil2003, Shea2004}.

Analysing community response to disturbance requiers a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and which reveals the pool of species fostered (or hampered) by
disturbance \citep{Lavorel2002, Bellwood2006}. Then, diversity metrics
encompassing both richness and evenness should also be considered to
assess the changes in community abundance distribution. Besides,
functional approaches have been shown to usefully complement pure
taxonomic approaches because they shed light on the species biological
attributes that directly link community diversity, composition and
redundancy to ecosystem functioning \citep{Violle2007b, Baraloto2012a}.
In that respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.

@@

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, (i) we draw taxonomic and functional post-disturbance
trajectories and examine the underlying ecological process, (ii) we
discuss the scope of the IDH regarding the different facets of community
diversity and the time after disturbance, and (iii) we analyze community
resilience and time to recovery.

@@

Paracou station in French Guiana (5\textdegree 18'N and
52\textdegree 53'W) is located in a lowland tropical rain forest in a
tropical wet climate with mean annual temperature of 26\textdegree C,
mean annual precipitation averaging 2980 mm.y\textsuperscript{-1} (30-y
period) and a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \citep{Wagner2011}. Elevation ranges from
5 to 50 m. Across all plots the topography mainly corresponds to
hilltops or hillsides, while bottomlands cover less than 1 \% of the
area. Plots are shallow ferralitic acrisols over a layer of transformed
saprolite with low permeability and lateral drainage. Soil conditions
are homogeneous, to the exception of the highest hilltops where the
thick surface allows a free vertical drainage
\citep{Gourlet-Fleury2004}.

@@

Our analysis highlighted the decoupling between functional and taxonomic
trajectories, with taxonomic trajectories maintaining initial
differences among plots while functional trajectories converged in the
functional space.

The humped-shaped taxonomic trajectories which amplitude varied with
disturbance intensity validated the IDH. Taxonomic diversity was
maximized at an intensity threshold of 20-25\% AGB loss and 20 years
after disturbance, when the recruitment of pioneers and
late-successionals were balanced. The IDH however did not apply to the
functional diversity. The decoupling between taxonomic and functional
response was linked to the disturbance impact on the functional
redundancy, that proved determining in community recovery.

@@

Regarding taxonomic composition, the different tree communities were
substantially different before disturbance, and this is visualized by
their distinct starting points on the NMDS axis 2. These initial
differences were maintained all along the 30 years following
disturbance, with the disturbance leading a displacement on the NMDS
axis 1 only. Changes were thus similar and may correspond to the
recruitment of a group of pioneers, e.g. \emph{Cecropia} spp.,
\emph{Vismia} spp., shared by all plots and whatever their initial
taxonomic differences and the intensity of disturbance
\citep{Denslow2000, Bongers2009}. Taxonomic trajectories seems to start
a recovery to the initial composition which, although far from fully
achieved after 30 years, suggested that taxonomic composition is
resilient and that the initial differences will be maintained in time
\citep{Folke2006}. This would show that species not belonging to the
pre-disturbance community were rarely recruited in the long-term,
probably because of the common dispersal limitations among tropical tree
species \citep{Svenning2005}.

Regarding functional composition, initial communities had similar
starting points and this contrasts to taxonomic composition. This means
that, despite differences in species composition, the functional
signature of the tree communities were quite similar. Following
disturbance, functional trajectories were consistent in the NMDS plane
with displacement intensities linked to disturbance intensities and thus
relied upon the recruitment of a shared pool of functional types
previously infrequent or absent. This common pool is composed of pioneer
resource-acquisitive (low leaf toughness, wood specific gravity, maximum
height and high Specific Leaf area) strategies that translated by a
displacement to the right along the first NMDS axis
\citep{Westoby1998, Wright2004, Reich2014}. Thereafter, the pioneers
recruited primarily were progressively excluded by long-lived, more
competitive and shade-tolerant species. The community functional
composition then quickly returned towards more resource-conservative
strategies, suggesting the recovery of the initial functional
composition. This recovery translated in the functional plane by a
displacement left along the first axis and upward along the second axis
(Fig. \ref{fig:NMDSplans}).

Both taxonomic and functional composition trajectories initiated a
return towards pre-disturbance state after 30 years, which highlighted
the taxonomic and functional resilience of communities. Taxonomic and
functional trajectories however appeared decoupled: while taxonomic
trajectories maintained the initial differences among communities, the
functional trajectories where similar and convergent in the functional
space \citep{Fukami2005} exemplifying the simultaneous operation of
trait-based assembly rules and species-level priority effects that
shapes tree community assembly in this forest, making it both
deterministic in functional space and historically contingent in
taxonomic space.

@@

Trajectories of taxonomic richness and evenness were markedly different
below and above an intensity threshold (\ref{fig:DivTaxo}). Below the
intensity threshold, for disturbances in the order of treatments 1 and
2, both taxonomic richness and evenness increased according to the
disturbance intensity \citep{Martin2015, Chaudhary2016}. This suggested
that while trees surviving after disturbance remained numerous enough to
maintain the richness of the pre-disturbance community
\citep{Bongers2009}, the recruitment of pioneers previously infrequent
or absent allowed increasing the taxonomic richness
\citep{Martin2015, Chaudhary2016}. As these pioneers became more
dominant, they balanced the usual hyper-dominance of tropical forests
and increased the taxonomic evenness \citep{Baraloto2012a}. Above the
intensity threshold, like for treatment 3, the taxonomic richness did
not exceed the initial value. The richness of surviving trees was then
too low to be offset by the recruitment of pioneers. In the Guiana
shield indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}.

Considering different moments of the post-disturbance trajectories,
there was always a humped-shaped pattern linking taxonomic richness or
evenness and the intensity of disturbance (\ref{fig:IDHplot}). Both
taxonomic richness and evenness were maximized at an intermediate
intensity, around 20-25\% of AGB lost, and this along the 30 years
following disturbance.

Regarding community functional trajectories, no marked differences were
observed among the post-disturbance trajectories (\ref{fig:DivTaxo}).
Whenever the disturbance intensity, functional richness and evenness
first increased with the recruitment of pioneers that were functionally
highly different from the pre-disturbance community
\citep{Denslow1980, Molino2001}. Then, after 15 to 20 years community
functional richness and evenness started to decrease, with the decline
of the first established species. Right after disturbance, short-lived
pioneers benefited from the resources made available and prevented the
establishment of other species. These pioneers then declined, which
decreased the functional richness and evenness \citep{Walker2009}.

Considering different times after disturbance, both functional richness
and evenness kept increasing with the disturbance intensity and did not
display humped-shaped trajectories (\ref{fig:IDHplot}).

@@

Post-disturbance trajectories of tree community composition and
diversity were both shaped by the recruitment of a determined pool of
pioneers identical among local communities and independent of the
disturbance intensity. (i) Composition trajectories in taxonomic and
functional space nevertheless appeared decoupled. While functional
trajectories remained similar in the functional space and converged
towards the recovery of a comparable initial state, taxonomic
trajectories showed initial differences in community composition that
were maintained along time. Community high functional redundancy
mediated this decoupling, as the loss of a species does not necessarily
entails the loss of its functional charactristics. (ii) Diversity
trajectories were contrasted as well. While the functional trajectories
remained similar whatever the disturbance intensity, taxonomic
trajectories were markedly different with a threshold at 20-25\% AGB
removed for which taxonomic richness was maximized. The Intermediate
Disturbance Hypothesis applied well to taxonomic diversity, but not to
the functional diversity. Taxonomic diversity was maximized at an
intermediate time after disturbance, that was around 25 years after
disturbance when the recruitment of early- and late-successional species
were balanced. Whenever the disturbance intensity, community resilience
(in terms of recovery of the pre-disturbance state) was tangible but
required several decades and relied upon the random lottery recruitment
of rare species. Given the long-term impacts of disturbance observed, we
suggest that 30 years is not enough time for tropical communities to
recover, even after relatively low intensity disturbance. Much of
community response to disturbance rely on species recruitment: refined
understanding of post-disturbance trajectories would be given by a
closer analysis of recruitment processes.

>>>>

% User-adder preamble
\usepackage{textcomp} \usepackage{tabu} \usepackage{caption}
\captionsetup{justification = justified}
\renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
\hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}

@@

\Abstract{
Anticipate the impact of the global change context on tropical forests
is now urgent. This calls to highlight the taxonomic and functional
facets of tree community post-disturbance trajectories and the
underlying processes. It also calls to clarify the link between
biodiversity and disturbance, specifically the scope of the Intermediate
Disturbance Hypothesis (IDH) that is debated in tropical forests. In
this study, we analyzed the tree community trajectories following a
disturbance gradient in a Neotropical forest over 30 years. We
considered community taxonomic and functional diversity, composition and
redundancy trajectories. We based on the annual botanical inventories of
75 ha of a Neotropical forest and on large trait datasets comprising 7
leaf, stem and life-history traits. We highlighted a decoupling between
taxonomic composition trajectories, that differed among communities, and
functional trajectories, that remained similar and convergent. The
taxonomic diversity followed humped-shaped trajectories depending on the
disturbance intensity that validated the IDH (Intermediate Disturbance
Hypothesis), while functional diversity trajectories homogeneous among
all plots dismissed the IDH. We explained this decoupling by the
variations in community functional redundancy that mitigates the
functional impact of disturbance. Although consistent, the recovery of
community composition, diversity and redundancy remained unachieved
after 30 years. This acknowledged the need of decades-long cycles
without disturbance to ensure a complete recovery, and questioned
tropical forest community resilience after repeated disturbances.
}

@@

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events, stating that community diversity peaks
at intermediate level of disturbance \citep{Connell1978}. The theory
assumes that the environmental fluctuations following disturbance foster
both competitively superior species and fast colonizers, thus preventing
competitive exclusion \citep{Shea2004, Pulsford2016}. In tropical
forests, however, observations of the IDH often diverge from theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. In this context it is urgent to
test the IDH and determine its scope, as the underlying processes might
be complicated by the huge diversity of tropical tree communities
\citep{Lindenmayer2012, Garcia_florez2017}.

Analysing community response to disturbance requiers a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and reveals the pool of species fostered (or hampered) by disturbance
\citep{Lavorel2002, Bellwood2006}. Then, diversity metrics encompassing
both richness and evenness should also be considered to assess the
changes in community abundance distribution. Besides, functional
approaches have been shown to usefully complement pure taxonomic
approaches as they shed light on the species biological attributes that
directly link community diversity, composition and redundancy to
ecosystem functioning \citep{Violle7b, Baraloto2012a}. In that respect,
a vast literature allowed recognizing major traits that represent
species ecological strategy and determine how they respond to changing
conditions \citep{Diaz2005}. Specifically in tropical forests, the
functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.

@@

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, (i) we draw taxonomic and functional post-disturbance
trajectories and examine the underlying ecological process, (ii) we
discuss the scope of the IDH regarding taxonomic and functional facets
of community diversity, and (iii) we analyze community resilience and
time to recovery.

@@

Paracou station in French Guiana (5° 18'N and 52° 53'W) is located in a
lowland tropical rain forest in a tropical wet climate with mean annual
temperature of 26° C, mean annual precipitation averaging 2980
mm.y\textsuperscript{-1} (30-y period) and a 3-month dry season
(\textless{} 100 mm.month\textsuperscript{-1}) from mid-August to
mid-November, and a one-month dry season in March \citep{Wagner2011}.
Elevation ranges from 5 to 50 m. Across all plots the topography mainly
corresponds to hilltops or hillsides, while bottomlands cover less than
1 \% of the area. Plots are shallow ferralitic acrisols over a layer of
transformed saprolite with low permeability and lateral drainage. Soil
conditions are homogeneous, to the exception of the highest hilltops
where the thick surface allows a free vertical drainage
\citep{Gourlet-Fleury2004}.

@@

Our analysis highlighted the decoupling between functional and taxonomic
trajectories. In terms of composition, while taxonomic trajectories
maintained the initial differences among plots, the functional
trajectories converged in the functional space. In terms of diversity,
only the taxonomic trajectories validated the IDH that explained the
humped-shaped taxonomic trajectories which amplitude depended on the
disturbance intensity. The decoupling between taxonomic and functional
response was explained by variations of community functional redundancy,
that mitigated the functional impact of disturbance and appeared as a
determinant of community recovery.

@@

Community taxonomic composition substantially differed before
disturbance, as materialized by their distinct location on the NMDS axis
2. These initial differences were maintained all along the 30 years
following disturbance, with the disturbance leading a displacement on
the NMDS axis 1 only. Taxonomic composition changes were similar among
plot and may correspond to the recruitment of a group of pioneers, e.g.
\emph{Cecropia} spp., \emph{Vismia} spp., shared by all plots whatever
their initial taxonomic differences and the intensity of disturbance
\citep{Denslow2000, Bongers2009}. Taxonomic trajectories initiated a
recovery towards the initial composition which, although far from being
achieved after 30 years, suggested the resilience of community taxonomic
composition and the maintenance of initial composition differences along
time \citep{Folke2006}. This suggested that species not belonging to the
pre-disturbance community were rarely recruited in the long-term,
probably because of the common dispersal limitations among tropical tree
species \citep{Svenning2005}.

Community functional composition in contrast followed similar
trajectories in the functional space and the amplitude of the
compositional changes depended on the disturbance intensity. Functional
trajectories were probably driven by the recruitment of a pool of
species common to all plots that were infrequent or absent before
disturbance. This common pool was composed of pioneers with
resource-acquisitive strategies (low leaf toughness, wood specific
gravity, maximum height and high Specific Leaf area) which recruitment
drove a displacement to the right along the first NMDS axis
\citep{Westoby1998, Wright2004, Reich2014}. Thereafter, the first
recruited pioneers were progressively excluded by long-lived, more
competitive and shade-tolerant species, which suggested the recovery of
the initial functional composition. This recovery translated into a
return towards resource-conservative strategies and in the functional
plane by a displacement left along the first axis and upward along the
second axis (Fig. \ref{fig:NMDSplans}).

The return after 30 years towards pre-disturbance state of composition
trajectories highlighted the taxonomic and functional resilience of
communities. Taxonomic and functional trajectories however appeared
decoupled: while taxonomic trajectories maintained the initial
differences among communities, the functional trajectories where similar
and convergent in the functional space \citep{Fukami2005}. This
decoupling exemplified the simultaneous operation of trait-based
assembly rules and species-level priority effects that shapes tree
community assembly in this forest, making it both deterministic in
functional space and historically contingent in taxonomic space.

@@

Trajectories of taxonomic richness and evenness differed markedly below
and above an intensity threshold (\ref{fig:DivTaxo}). For disturbances
in the order of treatments 1 and 2, below the intensity threshold, both
taxonomic richness and evenness increased according to the disturbance
intensity \citep{Martin2015, Chaudhary2016}. This suggested that trees
surviving after disturbance remained numerous enough to maintain the
richness of the pre-disturbance community \citep{Bongers2009}, and that
the recruitment of pioneers previously infrequent or absent increased
the taxonomic richness \citep{Martin2015, Chaudhary2016}. The pioneers
thus recruited became more dominant, which balanced the usual
hyper-dominance of tropical forests and increased the taxonomic evenness
\citep{Baraloto2012a}. Above the intensity threshold, like for treatment
3, the taxonomic richness did not exceed the initial value. This
suggested that the richness of surviving trees was lower than this of
the pre-disturbance community,and the difference was not offset by the
recruitment of pioneers. In the Guiana shield indeed, the pool of true
pioneers specifically recruited after disturbance is restricted to a few
common genera (e.g. \emph{Cecropia} spp., \emph{Vismia} spp.)
\citep{Guitet2018}.

A humped-shaped pattern always linked both taxonomic richness and
evenness with the disturbance intensity, at different time after
disturbance (\ref{fig:IDHplot}). Both taxonomic richness and evenness
were maximized at an intermediate intensity, around 20-25\% of AGB lost.

Regarding community functional trajectories, no marked differences were
observed among post-disturbance trajectories (\ref{fig:DivTaxo}). First,
whenever the disturbance intensity, the recruitment of pioneers that
were functionally highly different from the pre-disturbance community
increased both community functional richness and evenness
\citep{Denslow1980, Molino2001} and hampered the establishment of other
species. Then, after 15 to 20 years, the first recruits declined and
were replaced by species functionally more similar to the
pre-disturbance community, which decreased the functional richness and
evenness \citep{Walker2009}.

@@

Post-disturbance trajectories of tree community composition and
diversity were driven by the recruitment of a determined pool of
pioneers, identical among local communities and independent of the
disturbance intensity. The taxonomic composition trajectories maintained
the initial differences among communities, while the functional
trajectories were similar and converged in the functoinal space towards
the recovery of the initial composition. The diversity trajectories were
contrasted as well. While the functional trajectories remained similar
whatever the disturbance intensity, taxonomic trajectories were markedly
different from a threshold of 20-25\% AGB removed that maximized the
taxonomic richness. The Intermediate Disturbance Hypothesis applied well
to taxonomic diversity, but not to the functional diversity. The
decoupling between taxonomic and functional trajectories was mediated by
the variations in functional redundancy, as the loss of a species does
not necessarily entails the loss of its functional charactristics.
Community resilience (in terms of recovery of the pre-disturbance state)
was tangible but required several decades and relied upon the random
lottery recruitment of rare species. Given the long-term impacts of
disturbance observed, we suggest that 30 years is not enough time for
tropical communities to recover, even after relatively low intensity
disturbance. Much of community response to disturbance rely on species
recruitment: refined understanding of post-disturbance trajectories
would be given by a closer analysis of recruitment processes.
